---
title: "Data Exploration for capstone-project-Beiwko"
author: Benon Okwi
  - name: "Benon Okwi"
    orcid: "0009-0005-4173-3395"
    email: "beok8551l@colorado.edu"
    affiliation:
      - name: "University of Colorado Boulder"
        department: "Department of Civil, Environmental and Architectural Engineering"
        city: "Boulder"
        state: "CO"
        country: "USA"
date: Jul 3, 2026
format: html
editor: visual
---

# Introduction

My project is on rural electrification in the world and the data is the percentage of rural access, taken from 1960 to 2025. The data is from the world bank and it contains the different countries and their country codes with the indicator names and the years the data was captured.

# Methods

Compiled from nationally representative household survey data, and occasionally census data, from sources going back as far as 1990 by the world bank Global Electrification Database (GED)

## Reading the Data

```{r}

library(tidyverse)

my_project_data <- read_csv(here::here("data/raw/Access_to_Electricity_Rural.csv"))

```

## Data Exploration Approach

## Initial Data Tidying

# Results

## Initial Data Exploration

```{r}

head(my_project_data)

dim(my_project_data)
```

# Conclusions

## Summary of Findings

## Questions and Next Steps
