---
title: "Data Exploration for capstone-project-Beiwko"
author:
  - name: "Benon Okwi"
    orcid: "0009-0005-4173-3395"
    email: "beok8551l@colorado.edu"
    affiliation:
      - name: "University of Colorado Boulder"
        department: "Department of Civil, Environmental and Architectural Engineering"
        city: "Boulder"
        state: "CO"
        country: "USA"
date: today
toc: true
format: html
execute:
  warning: false
editor: visual
---

# Introduction

My project is on rural electrification in the world and the data is the percentage of rural access, taken from 1960 to 2025(Populated years from 1990-2023). The data is from the world bank and it contains the different countries and their codes with the indicator names and their codes and the years the data was captured.

# Methods

Compiled from nationally representative household survey data, and occasionally census data, from sources going back as far as 1990 by the world bank Global Electrification Database (GED)

## Reading the Data

```{r}

library(tidyverse)
library(dplyr)
library(ggplot2)

my_project_data <- read_csv(
  here::here("data/raw/Access_to_Electricity_Rural.csv"),
  skip = 4)

countries_meta <- read_csv(
  here::here("data/raw/Metadata_country.csv"))
```

## Data Exploration Approach

## Initial Data Tidying

```{r}

tidy_data <- my_project_data |>
  pivot_longer(
    cols = `1960`:`2025`,
    names_to = "year",
    values_to = "access_rural_pct") |>
  mutate(year = as.integer(year)) |>
  filter(!is.na(access_rural_pct))


tidy_countries <- tidy_data |>
  left_join(countries_meta, by = "Country Code") |>
  filter(!is.na(Region))

write_csv(tidy_data, here::here("data/processed/tidy_data.csv"))

tidy_data <- tidy_data |>
  mutate(`Country Name` = str_trim(`Country Name`))
```

# Results

To focus on the East African Region (Uganda, Kenya, Tanzania, Ethiopia and Rwanda). Rural Electrification in the region was minimal to bumpy before 2000 upto mid-2000s where we see a signifcant take-off. Uganda saw meagre rural electrification prior to 2000 with below 2% of the population having access. There we tiny increaments in access after 2000, between 2000 and 2018, there were fluctuations recorded in access, with more significant rises recorded from 2018 to 2023. Kenya went from 7% in 2000 to 65% in 2022, they enjoyed more steaper increments in access across the whole region. Tanzania has a similar case with Uganda with insignicifant increments in access till around 2012 where access started to shoot up significantly through to 2022. Ethiopia shows gradual increase in access between 2004 and 2012 before experiencing more gradient increase in access post 2014. Signficant %access increase in Rwanda is recorded post 2010 having barely had any rural access prior to 2008.

## Initial Data Exploration

```{r}

head(tidy_data)

dim(tidy_data)

tidy_data |>
  group_by(`year`) |>
  summarise(mean = mean(access_rural_pct),
            median = median(access_rural_pct),
            sd = sd(access_rural_pct))



```

```{r}

tidy_data_yr_2022 <- tidy_data|>
  filter(year == 2022)
```

Data Visualization

@fig-eastafrica is a plot of % rural electrification access in East Africa vs Sub-Saharan Africa Averages from 2000-2022

```{r}



countries_of_interest <- c("Kenya", "Ethiopia", "Rwanda", "Tanzania", 
                           "Uganda", "Sub-Saharan Africa")

plot_data <- tidy_data |>
  filter(`Country Name` %in% countries_of_interest,
         year >= 2000, year <= 2022)

my_colors <- c(
  "Kenya" = "#2a78d6",
  "Ethiopia" = "#008300",
  "Rwanda" = "#1baf7a",
  "Tanzania" = "#eda100",
  "Uganda" = "#4a3aa7",
  "Sub-Saharan Africa" = "#898781")

my_linetypes <- c(
  "Kenya" = "solid",
  "Ethiopia" = "solid",
  "Rwanda" = "solid",
  "Tanzania" = "solid",
  "Uganda" = "solid",
  "Sub-Saharan Africa" = "dashed")

#| label: fig-eastafrica
#| fig-cap: "Rural electricity access, East Africa vs. Sub-Saharan Africa Averages, 2000-2022"
#| fig-alt: "Line chart showing rural electricity access percentage from 2000 to 2022. Kenya, Ethiopia, and Rwanda rise sharply from near zero to 40-65 percent, Uganda is up and down till 2015 when it takes a sharper rise"


ggplot(plot_data, aes(x = year, y = access_rural_pct, 
                      color = `Country Name`, linetype = `Country Name`)) +
  geom_line(linewidth = 1) +
  geom_point(size = 1.5) +
  scale_color_manual(values = my_colors) +
  scale_linetype_manual(
    values = my_linetypes, guide = "none") +
  labs(
    title = "Rural electricity access: East Africa vs Sub-Saharan Africa Averages (2000-2022)",
    x = "Year",
    y = "% of rural population with access",
    color = NULL
  ) +
  theme_minimal(base_size = 12) +
  theme(
    legend.position = "bottom",
    panel.grid.minor = element_blank(),
    plot.title = element_text(face = "bold"))

ggsave(here::here("output/rural_electricity_trend.png"), 
       width = 8, height = 5, dpi = 300)


```

@fig-region2022 shows rural electricity access by region and income group, 2022

```{r}

#| label: fig-region2022
#| fig-cap: "Rural electricity access by region and income group, 2022"
#| fig-alt: "Horizontal bar chart of rural electricity access by region in 2022. North America, Europe and Central Asia, and high-income countries are at or near 100 percent. Sub-Saharan Africa and low-income countries lag far behind at 25 to 31 percent, well below the world average of 84 percent."

region_data <- tidy_data |>
  filter(
    `Country Name` %in% c("North America", "Europe & Central Asia", "High income",
                          "Upper middle income", "South Asia", 
                          "Latin America & Caribbean", "East Asia & Pacific",
                          "Lower middle income", "World", 
                          "Sub-Saharan Africa", "Low income"),
    year == 2022)

ggplot(region_data, aes(x = access_rural_pct, 
                        y = fct_reorder(`Country Name`, access_rural_pct))) +
  geom_col(fill = "#2a78d6") +                    
  labs(title = "Rural Electricity Access by Region and Income Group in 2022", 
    x = "% of rural population with access",
    y = "Regions and Income Groups"
  ) +
  theme_minimal(base_size = 12) +
  theme(panel.grid.major.y = element_blank())
```

# Conclusions

\- Rural electricity access rose globally from \~67% to \~84% between 2000 and 2022, but progress was highly uneven across regions. - East African countries (Kenya, Ethiopia, Rwanda) saw the largest gains, several multiplying access five- to twenty-fold from very low bases(see @fig-eastafrica). - Nigeria, despite a higher 2000 starting point, saw almost no sustained improvement and was overtaken by countries that started far behind it. - Sub-Saharan Africa and low-income countries overall lag significantly, with 40 countries still below 50% rural access in 2022. - Closing the remaining gap likely requires targeted national policy, not just general economic growth.
