---
title: "My Capstone Project"
author:
  - name: "Benon Okwi"
    orcid: "0009-0005-4173-3395"
    email: "beok8551l@colorado.edu"
    affiliation:
      - name: "University of Colorado Boulder"
        department: "Department of Civil, Environmental and Architectural Engineering"
        city: "Boulder"
        state: "CO"
        country: "USA"
date: today
format: html
toc: true
execute:
  warning: false
editor: visual
---

# Introduction

My project is on rural electrification in the world and the data is the percentage of rural access, taken from 1960 to 2025(Populated years from 1990-2023). The data is from the world bank and it contains the different countries and their codes with the indicator names and their codes and the years the data was captured.

# Methods

Compiled from nationally representative household survey data, and occasionally census data, from sources going back as far as 1990 by the world bank Global Electrification Database (GED) The data is initially in wide format with one country per row and one year per column

## Reading the Data

we use the skip = 4 to skip the four metadata lines that appear above the actual column headers

```{r}

library(tidyverse)
library(dplyr)
library(ggplot2)
library(gt)

my_project_data <- read_csv(
  here::here("data/raw/Access_to_Electricity_Rural.csv"),
  skip = 4)

countries_meta <- read_csv(
  here::here("data/raw/Metadata_country.csv"))
```

## Data Exploration Approach

## Initial Data Tidying

We reshape the data to long format using pivot_longer() and we filter out the non numeric columns. We also filter out the non numeric data in the regions.

```{r}

tidy_data <- my_project_data |>
  pivot_longer(
    cols = `1960`:`2025`,
    names_to = "year",
    values_to = "access_rural_pct") |>
  mutate(year = as.integer(year)) |>
  filter(!is.na(access_rural_pct))


tidy_countries <- tidy_data |>
  left_join(countries_meta, by = "Country Code") |>
  filter(!is.na(Region))

write_csv(tidy_data, here::here("data/processed/tidy_data.csv"))

tidy_data <- tidy_data |>
  mutate(`Country Name` = str_trim(`Country Name`))
```

# Results

To focus on the East African Region (Uganda, Kenya, Tanzania, Ethiopia and Rwanda). Rural Electrification in the region was minimal to bumpy before 2000 upto mid-2000s where we see a signifcant take-off. Uganda saw meagre rural electrification prior to 2000 with below 2% of the population having access. There we tiny increaments in access after 2000, between 2000 and 2018, there were fluctuations recorded in access, with more significant rises recorded from 2018 to 2023. Kenya went from 7% in 2000 to 65% in 2022, they enjoyed more steaper increments in access across the whole region. Tanzania has a similar case with Uganda with insignicifant increments in access till around 2012 where access started to shoot up significantly through to 2022. Ethiopia shows gradual increase in access between 2004 and 2012 before experiencing more gradient increase in access post 2014. Signficant %access increase in Rwanda is recorded post 2010 having barely had any rural access prior to 2008.

## Initial Data Exploration

```{r}

head(tidy_data)

dim(tidy_data)

tidy_data |>
  group_by(`year`) |>
  summarise(mean = mean(access_rural_pct),
            median = median(access_rural_pct),
            sd = sd(access_rural_pct))



```

```{r}

tidy_data_yr_2022 <- tidy_data|>
  filter(year == 2022)
```

## **Data Visualization**

@fig-eastafrica is a plot of % rural electrification access in East Africa vs Sub-Saharan Africa Averages from 2000-2022. It shows how access has grown in the region over the last 22 years with a mix of bumps and gradual increase in access.

```{r}
#| label: fig-eastafrica
#| fig-cap: "Rural electricity access, East Africa vs. Sub-Saharan Africa Averages, 2000-2022"
#| fig-alt: "Line chart showing rural electricity access percentage from 2000 to 2022. Kenya, Ethiopia, and Rwanda rise sharply from near zero to 40-65 percent, Uganda is up and down till 2015 when it takes a sharper rise"

countries_of_interest <- c("Kenya", "Ethiopia", "Rwanda", "Tanzania", 
                           "Uganda", "Sub-Saharan Africa")

plot_data <- tidy_data |>
  filter(`Country Name` %in% countries_of_interest,
         year >= 2000, year <= 2022)

my_colors <- c(
  "Kenya" = "#2a78d6",
  "Ethiopia" = "#008300",
  "Rwanda" = "#1baf7a",
  "Tanzania" = "#eda100",
  "Uganda" = "#4a3aa7",
  "Sub-Saharan Africa" = "#898781")

my_linetypes <- c(
  "Kenya" = "solid",
  "Ethiopia" = "solid",
  "Rwanda" = "solid",
  "Tanzania" = "solid",
  "Uganda" = "solid",
  "Sub-Saharan Africa" = "dashed")


ggplot(plot_data, aes(x = year, y = access_rural_pct, 
                      color = `Country Name`, linetype = `Country Name`)) +
  geom_line(linewidth = 1) +
  geom_point(size = 1.5) +
  scale_color_manual(values = my_colors) +
  scale_linetype_manual(
    values = my_linetypes, guide = "none") +
  labs(
    title = "Rural electricity access: East Africa vs Sub-Saharan Africa Averages (2000-2022)",
    x = "Year",
    y = "% of rural population with access",
    color = NULL
  ) +
  theme_minimal(base_size = 12) +
  theme(
    legend.position = "bottom",
    panel.grid.minor = element_blank(),
    plot.title = element_text(face = "bold"))

ggsave(here::here("output/rural_electricity_trend.png"), 
       width = 8, height = 5, dpi = 300)


```

@fig-region2022 shows rural electricity access by region and income group, 2022. Notably, North America, Europe and Central Asia are doing wonderfully similar to High-income countries with Sub-Saharan Africa and Low-Income countries lagging behind.

```{r}
#| label: fig-region2022
#| fig-cap: "Rural electricity access by region and income group, 2022"
#| fig-alt: "Horizontal bar chart of rural electricity access by region in 2022. North America, Europe and Central Asia, and high-income countries are at or near 100 percent. Sub-Saharan Africa and low-income countries lag far behind at 25 to 31 percent, well below the world average of 84 percent."

region_data <- tidy_data |>
  filter(
    `Country Name` %in% c("North America", "Europe & Central Asia", "High income",
                          "Upper middle income", "South Asia", 
                          "Latin America & Caribbean", "East Asia & Pacific",
                          "Lower middle income", "World", 
                          "Sub-Saharan Africa", "Low income"),
    year == 2022)

ggplot(region_data, aes(x = access_rural_pct, 
                        y = fct_reorder(`Country Name`, access_rural_pct))) +
  geom_col(fill = "#2a78d6") +                    
  labs(title = "Rural Electricity Access by Region and Income Group in 2022", 
    x = "% of rural population with access",
    y = "Regions and Income Groups"
  ) +
  theme_minimal(base_size = 12) +
  theme(panel.grid.major.y = element_blank())
```

@tbl-summary shows summary statistics for rural electrification Access in East Africa and Sub-Saharan Region, 2000 - 2022 with Kenya having the highest mean % and Maximum in the region and Tanzania statistics showing there is very slow and minimal access. The East African region is below the Sub-Saharan region averages bar Kenya.

```{r}
#| label: tbl-summary
#| tbl-cap: "Summary statistics for rural electricity access in East African Countries, 2000 - 2022"

summary_table <- plot_data |>
  group_by(`Country Name`) |>
  summarise(
    Mean = mean(access_rural_pct, na.rm = TRUE),
    Median = median(access_rural_pct, na.rm = TRUE),
    SD = sd(access_rural_pct, na.rm = TRUE),
    Min = min(access_rural_pct, na.rm = TRUE),
    Max = max(access_rural_pct, na.rm = TRUE),
    N = n()
  ) |>
  arrange(desc(Mean))

summary_table |>
  gt() |>
  cols_label(
    `Country Name` = "Country",
    Mean = "Mean (%)",
    Median = "Median (%)",
    SD = "Std. dev.",
    Min = "Min (%)",
    Max = "Max (%)",
    N = "Years"
  ) |>
  fmt_number(columns = c(Mean, Median, SD, Min, Max), decimals = 1) |>
  cols_align(align = "center", columns = c(Mean, Median, SD, Min, Max, N))

```

# Conclusions

- Rural electricity access rose globally from \~67% to \~84% between 2000 and 2022, but progress was highly uneven across regions.

- East African countries (Kenya, Ethiopia, Rwanda) saw the largest gains, several multiplying access five- to twenty-fold from very low bases(see @fig-eastafrica). - Nigeria, despite a higher 2000 starting point, saw almost no sustained improvement and was overtaken by countries that started far behind it.

- Sub-Saharan Africa and low-income countries overall lag significantly, with 40 countries still below 50% rural access in 2022.

-  Closing the remaining gap likely requires targeted national policy, not just general economic growth.

## **Summary of Findings** 

Since my reason for choosing rural electrification was to to see how the progress has been for rural Uganda where i come from, I tried to focus my visualizations on Uganda and East Africa. The trajectory of % access in uganda was generally underwhelming for the most part. In the last 5 years the government rolled out a Rural Electrification project so the next World Bank data post 2025 will surely and hopefully show a more exponential growth curve in access.
