---
title: "Data Exploration for capstone-project-Beiwko"
author:
  - name: "Benon Okwi"
    orcid: "0009-0005-4173-3395"
    email: "beok8551l@colorado.edu"
    affiliation:
      - name: "University of Colorado Boulder"
        department: "Department of Civil, Environmental and Architectural Engineering"
        city: "Boulder"
        state: "CO"
        country: "USA"
date: today
format: html
editor: visual
---

# Introduction

My project is on rural electrification in the world and the data is the percentage of rural access, taken from 1960 to 2025(Populated years from 1990-2023). The data is from the world bank and it contains the different countries and their codes with the indicator names and their codes and the years the data was captured.

# Methods

Compiled from nationally representative household survey data, and occasionally census data, from sources going back as far as 1990 by the world bank Global Electrification Database (GED)

## Reading the Data

```{r}

library(tidyverse)
library(dplyr)

my_project_data <- read_csv(
  here::here("data/raw/Access_to_Electricity_Rural.csv"),
  skip = 4)

countries_meta <- read_csv(
  here::here("data/raw/Metadata_country.csv"))
```

## Data Exploration Approach

## Initial Data Tidying

```{r}

tidy_data <- my_project_data |>
  pivot_longer(
    cols = `1960`:`2025`,
    names_to = "year",
    values_to = "access_rural_pct") |>
  mutate(year = as.integer(year)) |>
  filter(!is.na(access_rural_pct))


tidy_countries <- tidy_data |>
  left_join(countries_meta, by = "Country Code") |>
  filter(!is.na(Region))

write_csv(tidy_data, here::here("data/processed/tidy_data.csv"))
```

# Results

## Initial Data Exploration

```{r}

head(tidy_data)

dim(tidy_data)

tidy_data |>
  group_by(`year`) |>
  summarise(mean = mean(access_rural_pct),
            median = median(access_rural_pct),
            sd = sd(access_rural_pct))



```

# Conclusions

## Summary of Findings

## Questions and Next Steps
